---
output: github_document 
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

```

# The `khsmisc` package

<!-- badges: start -->
<!-- badges: end -->

The `khsmisc` package serves two purposes: (a) loading a set of core packages for epidemiology research and (b) providing a set of custom functions, which also build on the core packages.

## Installation

The `khsmisc` package can be installed from [GitHub](https://github.com/stopsack/khsmisc). A CRAN submission is not planned.

```r
# if "remotes" library is missing, install it first:
#   install.packages("remotes")
library(remotes)
remotes::install_github("stopsack/khsmisc")
```


## Loaded Core Packages

When loading `library(khsmisc)`, by design, the following dependencies will be made available:

* **Data handling:** the whole universe of the `tidyverse`;
variable labels via `labelled`; reading Excel files via `readxl`.
* **Tables:** formatted tables via `gt`; descriptive tables via `gtsummary`.
* **Figures:** `ggplot2`; red/green blindness-proof color palettes via `viridis`;
  arranging figure panels via `cowplot`.
* **Analyses:** time-to-event analyses using `survival`;
  accessing regression model results via `broom`.

## Additional Functions for Epidemiology Research

* **Data handling:** `varlist` for variable list/dataset inventory, Stata style; `write_csv_safely`, `save_safely`, and `pdf_safely` to export CSV files, R objects, or PDFs without overwriting existing files.
* **Functions for vectors:** `scale_to_range` to scale a vector to a given range; `Mode` to obtain the most common value; `roundp` to round and format *p*-values
* **Functions for descriptive tables:** `mytab` for categorical variables; `tsummary` for continuous variables; `mygt` for formatting and printing any tibble as a 'gt' table; `table1` for formatted stratified descriptive tables using `gtsummary`; `rates` for event counts, person-time, incidence rates, and cumulative incidence.
* **Functions for figures:** `stripplot` for box/whiskers plots overlaid the a dot plot of all data points


# Sample analysis workflow

## Setting up 

First, install `khsmisc` [as described above](#installation). Then:

(@) **Create an RMarkdown file**

    RStudio: *File* > *New File* > *R Markdown...* > *Create Empty Document*

(@) **Add a header for the RMarkdown**

    Copy the following so-called "YAML" header into the empty document, including the three dashes (`---`) at the beginning and end:
    
```{yaml, eval = FALSE}
---
title: "A test project: Aneuploidy in TCGA-BLCA"
author: "Konrad H. Stopsack <stopsack@mskcc.org>"
date: "`r format(Sys.time(), '%b %d, %Y %X')`"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 2
    number_sections: true
    theme: cosmo
    toc_float:
      collapsed: false
      smooth_scroll: false
---
```

   + Change the `title:` and `author:` fields. 
   + The date will automatically be updated when compiling the document. 
   + Note that this type of code, "YAML," is sensitive to exact identation. Every indentation is done by exactly two spaces.

(@) **Load `khsmisc` library**
    
    Click *Insert* > *R* in the toolbar. Write into the new piece of code ("chunk"):

```r
library(khsmisc)
```

(@) **Knit the RMarkdown a first time**

    You are ready to compile the RMarkdown using the *Knit* button. An HTML document should open that has nothing but a title and the startup messages of `khsmisc` about all the packages it loads. If an an error message is shown instead, perhaps `khsmisc` was not installed properly? 
    
    Learn more about the syntax of RMarkdown in the [RStudio cheatsheets](http://www.rstudio.com/resources/cheatsheets/).


```{r quietload, include = FALSE, message = FALSE}
library(khsmisc)
```

## Data handling

### Load the TCGA `BLCA` dataset

Participant and tumor data from the Cancer Genome Atlas Bladder Cancer "cohort" (`BLCA`) will be used in their published 2017 version. Insert a new R chunk into the markdown (toolbar: *Insert* > *R*), and add code for reading a tab-separated file ("TSV"). It retrieves the file from the `data` directory and skips the first four lines.
     
```{r}
clinical <- read_tsv(file = "../data/data_clinical_patient.txt", 
                     skip = 4)
```

An inventory of the dataset:

```{r}
varlist(clinical) %>%
  print(n = 15)  # show the first 15 lines of output
```

Select and rename variables of interest:

```{r}
clinical <- clinical %>%
  select(id         = PATIENT_ID, 
         sex        = SEX,
         race       = RACE,
         ethnicity  = ETHNICITY,
         height     = HEIGHT,
         weight     = WEIGHT,
         smoke      = TOBACCO_SMOKING_HISTORY_INDICATOR,
         agedx      = AGE, 
         dxyear     = INITIAL_PATHOLOGIC_DX_YEAR,
         c_tstage   = CLIN_T_STAGE,
         p_stage    = AJCC_PATHOLOGIC_TUMOR_STAGE,
         p_tstage   = AJCC_TUMOR_PATHOLOGIC_PT, 
         p_nstage   = AJCC_NODES_PATHOLOGIC_PN,
         mstage     = AJCC_METASTASIS_PATHOLOGIC_PM,
         grade      = GRADE,
         histology  = HISTOLOGICAL_SUBTYPE,
         os_status  = OS_STATUS, 
         os_mos     = OS_MONTHS, 
         dfs_status = DFS_STATUS,
         dfs_mos    = DFS_MONTHS)
```

### Load Taylor *et al.* aneuploidy calls

Additional exposure data on tumors of the same study participants comes in the form of derived tumor aneuploidy scores as per Taylor, ..., Meyerson ([Cancer Cell 2018;33:676–689](https://doi.org/10.1016/j.ccell.2018.03.007)).

    
Load the Excel file provided as their Supplementary Table 2, show inventory, and select/rename variables:
    
```{r}
taylor <- read_xlsx("../data/Taylor2018-TableS2-1-s2.0-S1535610818301119-mmc2.xlsx", 
                    skip = 1)  # skip the first line

varlist(taylor) %>%
  print(n = 20)

taylor <- taylor %>%
  select(id         = Sample,
         ascore     = `AneuploidyScore(AS)`,
         ascore_del = AS_del,
         ascore_amp = AS_amp,
         doubling   = Genome_doublings,
         purity     = Purity)
```


### Merging datasets

Left-join the Taylor *et al.* aneuploidy scores to the clinical data, *i.e.*, keep all observations from `clinical` and those that match from `taylor`. Because IDs in `taylor` are tumor IDs, they contain an extra suffix that we need to remove before merging. We also need to check that this procedure did not introduce duplicates.
    
```{r}
taylor <- taylor %>%
  mutate(id = str_sub(string = id, start = 1, end = -4))

combined <- clinical %>%
  left_join(taylor, by = "id")

sum(duplicated(combined$id))  # Check for duplicates. Should return 0.
```
   

### Recoding variables

Consistently code categorical variables as a `factor` and continuous variables as `numeric`:

```{r}
combined <- combined %>%
  mutate_at(.vars = vars(sex, race, ethnicity, smoke,
                         c_tstage, p_stage, p_tstage, p_nstage, mstage, 
                         grade, histology, 
                         os_status, dfs_status),
            .funs = factor) %>%
  mutate_at(.vars = vars(height, weight, agedx, dxyear, os_mos, dfs_mos),
            .funs = as.numeric)
```

The warning messages are expected, as some numeric variables contain non-numeric values.


Next, recode various labels of missing data in categorical variables to a consistent missing type, using `fct_collapse`:

```{r}
# Tabulate some examples
combined %>% select(race, dfs_status, histology) %>%
  map(.f = table, useNA = "always")  # always show NA ("missing") levels

# Collapse all actually missings to R's missing
combined <- combined %>%
  mutate_if(.predicate = is.factor, 
            .funs = ~fct_collapse(., NULL = c("[Not Available]", 
                                              "[Discrepancy]", 
                                              "Indeterminate")))

# Revisit the examples after recoding
combined %>% select(race, dfs_status, histology) %>%
  map(.f = table, useNA = "always")  # always show NA ("missing") levels
```


Recode event indicators to the numeric values that `survival` functions require:

```{r}
combined <- combined %>%
  mutate(dfs_event = case_when(dfs_status == "Recurred/Progressed" ~ 1,
                               dfs_status == "DiseaseFree"         ~ 0),
         os_event  = case_when(os_status  == "DECEASED"            ~ 1,
                               os_status  == "LIVING"              ~ 0))
```


Recode race categories to make them more readable:

```{r}
# Before:
combined %>%
  count(race)

combined <- combined %>% 
         # Change from all uppercase:
  mutate(race      = factor(str_to_title(race)),
         ethnicity = factor(str_to_title(ethnicity)),
         # Make "Black" as short as other categories
         race = fct_recode(race, Black = "Black Or African American"),
         # Make "White" the reference category because of sample size
         race = fct_relevel(race, "White"))

# After:
combined %>%
  count(race)
```

Assign meaningful labels to the `smoke` variable, based on inspection of other smoking-related variables in the full dataset:

```{r}
combined <- combined %>%
  mutate(smoke = factor(smoke, levels = 1:5, 
                        labels = c("Never", "Current", 
                                   "Former, quit >15 y", "Former, quit <15 y", 
                                   "Former")),
         # Combine all "formers":
         smoke3 = fct_collapse(smoke, Former = c("Former", 
                                                 "Former, quit >15 y", 
                                                 "Former, quit <15 y")),
         # Change order of levels, starting with "never" as the reference:
         smoke3 = fct_relevel(smoke3, "Never", "Former", "Current"))
```


### Creating derived variables

```{r}
combined <- combined %>%
  mutate(bmidx = weight / ((height/100)^2)) %>%
  select(-height, -weight)
```

### Categorizing variables

Category boundaries are informed by results from code to be run below.

```{r}
combined <- combined %>%
  mutate(ascore_cat = cut(ascore, 
                          breaks = c(0, 5, 10, 20, max(ascore, na.rm = TRUE)), 
                          include.lowest = TRUE))
```

### Labeling variables

```{r}
combined <- combined %>%
  set_variable_labels(sex        = "Sex",
                      race       = "Self-reported race",
                      ethnicity  = "Ethnicity",
                      smoke      = "Smoking status at diangosis",
                      smoke3     = "Smoking status at diagnosis",
                      agedx      = "Age at diagnosis",
                      dxyear     = "Calendar year of initial diagnosis",
                      bmidx      = "Body mass index at diagnosis",
                      p_stage    = "AJCC pathologic stage",
                      mstage     = "Metastases at diagnosis",
                      grade      = "Histologic grade",
                      histology  = "Histologic subtype",
                      ascore     = "Aneuploidy score",
                      ascore_cat = "Aneuploidy score",
                      ascore_del = "Aneuploidy score: deletions",
                      ascore_amp = "Aneuploidy score: amplifications",
                      doubling   = "Whole genome doubling",
                      purity     = "DNA tumor purity by ABSOLUTE")
```


## Data description

### Contingency tables

```{r}
combined %>%
  select(smoke3, race) %>%
  mytab() %>%  # make contingency table
  mygt()  # format
```

### Table of distributional statistics

Exposure and confounders

```{r}
combined %>%
  select(race, agedx, bmidx, dxyear, purity) %>%
  tsummary(by = "race") %>%  # make descriptive table
  select(-mean, -sd, -sum) %>%  # remove undesired statistics
  mutate_if(.predicate = is.numeric, .funs = round, digits = 2) %>%  # round
  mygt()  # format
```

Outcome

```{r}
combined %>%
  select(starts_with("ascore")) %>%
  tsummary() %>%  # make descriptive table
  select(-mean, -sd, -sum) %>%  # remove undesired statistics
  mygt()  # format
```


### Plots of the outcome

```{r}
combined %>%
  ggplot(mapping = aes(x = ascore)) +
  geom_histogram(binwidth = 2) +  # each bar = 2 units of "ascore"
  theme_minimal_grid()
```

Is the outcome reasonably normally distributed? Show a quantile--quantile plot before and after log transformation.

```{r}
combined %>% 
  ggplot(aes(sample = ascore)) + 
  stat_qq() + stat_qq_line() +
  labs(y = "Sample, untransformed") +
  theme_minimal_grid()

combined %>% ggplot(aes(sample = log(ascore + 0.1))) + 
  stat_qq() + stat_qq_line() +
  labs(y = "Sample, log(x + 0.1)-transformed") +
  theme_minimal_grid()
```


## Main results

### Study population: Applying and documenting exclusions

Because the main exposure is self-reported race, we will have to exclude participants with missing race from the analytical population.

```{r}
# All participants:
nrow(combined)

# Exclude with missing race:
analytical <- combined %>%
  filter(!is.na(race)) %>%
  copy_labels_from(from = combined)

# Analytical population:
nrow(analytical)
```

Note that the dataset contains participants with tumors that were metastatic at diagnosis (M1):
```{r}
analytical %>%
  count(mstage)
```

In the analytical dataset, does everyone have an aneuploidy score?

```{r}
analytical %>%
  mutate(missing_ascore = is.na(ascore)) %>%
  select(missing_ascore, race) %>%
  mytab() %>%
  mygt()
```


### Table 1: Exposure-stratified characteristics of the study population

```{r}
analytical %>%
  select(race, ethnicity, sex, agedx, dxyear, smoke3, bmidx, 
         p_stage, mstage, grade, histology, purity, doubling) %>%
  table1(by = "race", digits = list(agedx  = c(1, 1), 
                                    dxyear = c(0, 0)))
```


### Figure 1: Box-whiskers/dot plots

```{r}
analytical %>%
  stripplot(x = "race", y = "ascore")
```


### Table 2: Stratified and regression analyses

```{r}
# Part 1: Stratified
analytical %>%
  select(ascore_cat, race) %>%
  table1(by = "race")

# Function to run linear regression model and return "tidy" results
run_lm <- function(data, outcome, exposure, covariates = "") {
  # Fit linear model
  fit <- lm(formula = as.formula(paste(outcome, "~", exposure, covariates)),
            data = data)
  
  fit %>%
    # Obtain coefficients and confidence intervals
    tidy(conf.int = TRUE) %>%
    # restrict to exposure variable only
    filter(str_detect(string = term, pattern = paste0("^", exposure))) %>%
    # Round estimates
    mutate_at(.vars = vars(estimate, conf.low, conf.high),
              .funs = ~trimws(format(round(., digits = 1), nsmall = 1))) %>%
    # Concatenate point estimate and confidence interval into one string
    transmute(exp = str_remove(string = term, pattern = exposure),
              res = paste0(estimate, " (", conf.low, " to ", conf.high, ")")) %>%
    # Add reference category in front
    bind_rows(tibble(exp = data %>% 
                       select(exp = one_of(exposure)) %>% 
                       pull(exp) %>% 
                       levels(.) %>%
                       pluck(1), 
                     res = "0 (reference)"), 
              .) %>%
    # Make result table horizontal
    pivot_wider(names_from = exp, values_from = res) %>%
    # Add a column with the numnber of observations in model
    # to detect inadvertent exclusions due to missing data
    mutate(Observations = fit %>% glance() %>% pull(nobs))
}

# Run different outcome regression models:
  # (1) Define covariate sets
tribble(~Model,                         ~confounders,
       "Model (1) Age/sex-adjusted",    "+ agedx + sex",
       "Model (2) Model 1 + smoking",   "+ agedx + sex + smoke3",
       "Model (3) Model 2 + histology", "+ agedx + sex + smoke3 + histology",
       "Model (4) Model 3 + grade",     "+ agedx + sex + smoke3 + histology + grade") %>%
  # (2) Execute run_lm() for each row of the above tibble
  mutate(res = map(.x = confounders,
                   .f = run_lm,
                   # data, outcome, and exposure are the same each time:
                   data = analytical, outcome = "ascore", exposure = "race")) %>%
  # (3) Unnest the list column with the results
  unnest(cols = res) %>%
  # (4) Remove the cryptic and duplicative "confounders" column
  select(-confounders) %>%
  # Format the table
  mygt()
```


### Table 3: Effect modification

Models are adjusted for age (and sex, unless sex is an effect modifier).

```{r}
tribble(
  ~effectm,    ~Stratum, ~subset,                                                     ~confounders,
  "Sex",       "Female", analytical %>% filter(sex == "Female"),                     "+ agedx",
  "Sex",       "Male",   analytical %>% filter(sex == "Male"),                       "+ agedx",
  "Histology", "Papillary",     analytical %>% filter(histology == "Papillary"),     "+ agedx + sex",
  "Histology", "Non-Papillary", analytical %>% filter(histology == "Non-Papillary"), "+ agedx + sex") %>%
  mutate(res = map2(.x = subset, .y = confounders,
                    .f = run_lm,
                    outcome = "ascore", exposure = "race")) %>%
  unnest(cols = res) %>%
  group_by(effectm) %>%
  select(-subset, -confounders) %>%
  mygt()
````


### Table 4: Sensitivity analyses

...
